In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-19 tex/sigmasfr/ms.tex
edits to section 3.2

For the most compact galaxy (J0905+5759, $r_e=0.013\arcsec$ or
100~pc), we also show the observed one-dimensional surface brightness
profile in the bottom-right panel of Figure~\ref{fig:profile}.  We
compare to the profiles of six stars in the same image, the best-fit
de Vaucouleurs model, and a de Vaucoulerus model with a larger
effective radius ($r_e=0.04\arcsec$, the native WFC3/UVIS pixel size,
which corresponds to a physical scale of 290~pc).  This comparison
illustrates that this galaxy, while only marginally resolved with an
effective radius that is $\sim20\%$ of the image FWHM, is clearly more
extended than a point source.

@@

\subsection{Photometry, Stellar Masses, and Star-formation rates}

@@

We estimate IR-based star-formation rates for the 25/29 galaxies with
WISE 12 or 22~$\mu$m detections by fitting \citet{cha01} templates to
their 12 and 22~$\mu$m fluxes.  For the 14/29 galaxies with 22~$\mu$m
detections, this yields SFRs that agree with those obtained from the
\citet{ruj12} method based on 24~$\mu$m luminosity (with a scatter of
0.05~dex).  We note that several authors have shown that the shape of
the IR SED for star-forming galaxies depends on the surface density of
star formation \citep[e.g.,][]{ruj11,elb11}, with more compact
starbursts having larger total-IR (8--1000~$\mu$m) to mid-IR
(8--24~$\mu$m) ratios.  The extreme $\sigmasfr$ values of our sample
imply large total-IR to mid-IR ratios, characteristic of the most
luminous galaxies in the local universe
\citep[e.g.,][]{cha01,dal02,rie09}.  For the 8/29 sources with SFRs in
the ULIRG regime ($\sfrir>200$~\msun~yr$^{-1}$), we would obtain SFRs
that are larger by 0.5~dex (compared to the values obtained above) if
we used the most luminous local templates.

We also estimate star-formation rates and stellar masses based on
stellar population fits to the $\lrest=0.1$--3~$\mu$m SEDs using the
method of \citet{mou11}.  For galaxies with
$\sfrir>50$~\msun~yr$^{-1}$, there is agreement between these UV-based
SFR estimates and $\sfrir$ with a scatter of 0.32~dex.  We find a
median extinction of $A_V=0.4$~mag for an SMC-based dust law.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{spectra.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  These spectra are
  dominated by light from the young stellar population and exhibit
  \oii~$\lambda3727$, H$\beta$~$\lambda4861$, and \oiii~$\lambda5007$
  emission lines.  The panels on the right zoom on the
  \mgii~$\lambda\lambda2796,2803$ spectral region in both wavelength
  and velocity space.  The spectrum in the bottom panel has sufficient
  spectral resolution ($\textnormal{FWHM}\approx8$~\kms) to resolve
  the intrinsic shape of the absorption-line profile, revealing that
  the gas near the centroid velocity has unity covering factor.}
\label{fig:spectra}
\end{figure}

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} %\\
%\includegraphics[angle=0,scale=.7]{sizemass_sigmasfr.ps}
%\caption{$\sigmasfr$--mass and size--mass relationships.}
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%
  contours).  We mark the empirical threshold for launching winds
  \citep[dotted line, $\sigmasfr=0.1$~\units;][]{hec02}, the
  90th-percentile starburst intensity limit from \citet{meu97} (dashed
  line, $\sigmasfr=45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit.  The sample of gas-rich mergers includes ULIRGs
  \citep{vei06}, Lyman break analogs with dominant central objects
  \citep{ove09}, and the local compact starburst Arp 220 \citep{sco97,
    ken98, rod08}.  The samples of star-forming galaxies are drawn
  from \citet{wuy11} at $0.02<z<0.2$ (solid black contour),
  $0.5<z<1.5$ (dashed red countour), and $1.5<z<2.5$ (dotted purple
  contour), and from \citet{law12} at $2.5<z<3.6$ (dash-dotted blue
  contour). }
\label{fig:sigmasfr}
\end{center}
\end{figure*}

\subsection{Optical Spectroscopy}

We present low-resolution $R=1000-3000$ spectroscopy from MMT/Blue
Channel and SDSS (J1506+5402), Magellan/MagE (J1341-0321), and
Keck/LRIS (J0905+5759) in Figure~\ref{fig:spectra.}.  We also
highlight the \mgii~$\lambda\lambda2796,2803$ absorption lines, which
are used use to measure outflow velocities.

For J0905+5759 we show the \mgii\ absorption-line profiles from
$R\approx37,000$ spectra obtained with Keck/HIRES.  The oscillator
strength of \mgii~$\lambda$2796 is twice that of \mgii~$\lambda$2803,
but the equivalent widths of the two lines and the shapes of the line
profiles are quite similar, indicating that both lines are optically
thick.  The fact that the absorption line profile approaches zero near
the velocity centroid $v=-2470$~\kms\ indicates that the outflowing
gas covers the entire continuum source at these velocities.

@@

\subsection{Momentum Flux from Supernovae and Stellar Winds}

\citet{meu97} argued for a limit near 45~\units (somewhat smaller for
a Chabrier IMF) based on the surface brightness of starbursts measured
using rest-frame UV, H$\alpha$, far-IR, and radio continuum emission,
such that only 10\% of starbursts exceeded this limit.  They argued
that the momentum flux from supernovae and stellar winds expected at
this threshold matches the mean central pressure measured by
\citet{hec90} for starburst-driven winds.

There are also relevant arguments about ram pressure driving of
high-velocity winds from \citet{hec11}.

\subsection{The Eddington Limit from Radiation Pressure on Dust Grains}

In models where gaseous outflows are driven by radiation pressure from
massive stars, estimates of the maximum $\sigmasfr$ where the outward
force from radiation exceeds the inward force of gravity are in the
$\sigmasfr\sim10^3$~\units\ range \citep[e.g.,][]{tho05,hop10}.

@@

The relative importance of momentum injection versus thermal heating
may vary as a function of galaxy mass \citep[e.g.,][]{hop12}, with
thermal heating being less efficient in galaxies with larger gas
densities and shorter cooling times.  Regarding momentum injection,
\citet{sha11} have argued that radiation pressure becomes more
important than ram pressure at large SFRs.  Massive star clusters with
large gas surface densities are the ideal launching point for
galactic-scale outflows driven by radiation pressure
\citep[e.g.,][]{mur11}, and the outflow velocities are expected to
scale with escape velocity of the most massive star clusters in a
galaxy.

\subsection{Outflows at the Escape Velocity}

The observed $\gtrsim1000$~\kms\ outflow velocities we observe correspond
approximately to the escape velocities for compact regions (i.e.,
massive star clusters) with the sizes and masses that we've measured.

\begin{equation}
v_{esc} = \sqrt{2GM_{cl}/r_{cl}} = 2074~\textnormal{km~s}^{-1} \\%\left({v_c\over200}\right)^{5/2} 
\left({M_{cl}\over10^{11} M_{\odot}}\right)^{1/2} 
\left({r_{cl}\over200~\textnormal{pc}}\right)^{-1/2}
\label{eq:vesc}
\end{equation}

@@

Connection to ellipticals.  Conceivable that they could grown in size
sufficiently by $z=0$?

@@

@@

@@

@@

@@

>>>>

For the most compact galaxy (J0905+5759, $r_e=0.013\arcsec$ or
100~pc), we also show the observed one-dimensional surface brightness
profile in the bottom-right panel of Figure~\ref{fig:profile}.  We
compare to the profiles of six stars in the same image, the best-fit
de Vaucouleurs model, and a de Vaucouleurs model with a larger
effective radius ($r_e=0.04\arcsec$, the native WFC3/UVIS pixel size,
which corresponds to a physical scale of 290~pc).  This comparison
illustrates that this galaxy, while only marginally resolved with an
effective radius that is $\sim20\%$ of the image FWHM, is clearly more
extended than a point source.

@@

%\subsection{Photometry, Star Formation Rates, and Stellar Masses}
\subsection{Star Formation Rates and Stellar Masses}

@@

We estimate IR-based star-formation rates for the 25/29 galaxies with
WISE 12 or 22~$\mu$m detections by fitting \citet{cha01} templates to
their 12 and 22~$\mu$m fluxes.  For the 14/29 galaxies with 22~$\mu$m
detections, this yields SFRs that agree with those obtained from the
\citet{ruj12} method based on 24~$\mu$m luminosity with a scatter of
0.05~dex.  We note that several authors have shown that the shape of
the IR SED for star-forming galaxies depends on the surface density of
star formation \citep[e.g.,][]{ruj11,elb11}, with more compact
starbursts having larger total-IR (8--1000~$\mu$m) to mid-IR
(8--24~$\mu$m) ratios.  The extreme $\sigmasfr$ values of our sample
imply large total-IR to mid-IR ratios, characteristic of the most
luminous galaxies in the local universe
\citep[e.g.,][]{cha01,dal02,rie09}.  If we used the most luminous
local templates for the 8/29 sources with SFRs in the ULIRG regime
($\sfrir>200$~\msun~yr$^{-1}$), we would obtain SFRs that are larger
by 0.5~dex than the values we adopt for this paper.

We also estimate star-formation rates and stellar masses based on
stellar population fits to the $\lrest=0.1$--3~$\mu$m SEDs using the
method of \citet{mou11}.  For galaxies with
$\sfrir>50$~\msun~yr$^{-1}$, there is agreement between these UV-based
SFR estimates and $\sfrir$ with a scatter of 0.32~dex.  For an SMC
dust law, we find a median attenuation of $A_V=0.4$~mag.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{spectra.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  These spectra are
  dominated by light from the young stellar population and exhibit
  \oii~$\lambda3727$, H$\beta$~$\lambda4861$, and \oiii~$\lambda5007$
  emission lines.  The panels on the right highlight the
  \mgii~$\lambda\lambda2796,2803$ spectral region in both wavelength
  and velocity space to illustrate the outflow kinematics.  The
  spectrum in the bottom panel has sufficient spectral resolution
  ($\textnormal{FWHM}\approx8$~\kms) to resolve the intrinsic shape of
  the absorption-line profile, revealing that the gas near the
  centroid velocity ($v=-2470$~\kms) covers the entire galaxy.}
\label{fig:spectra}
\end{figure}

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} %\\
%\includegraphics[angle=0,scale=.7]{sizemass_sigmasfr.ps}
%\caption{$\sigmasfr$--mass and size--mass relationships.}
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%,
  95\%, and 99.7\% contours).  We mark the empirical threshold for
  launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit, suggesting growth that is limited by momentum
  injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}

% The sample of gas-rich mergers includes ULIRGs
%   \citep{vei06}, Lyman break analogs with dominant central objects
%   \citep{ove09}, and the local compact starburst Arp 220 \citep{sco97,
%     ken98, rod08}.  The samples of star-forming galaxies are drawn
%   from \citet{wuy11} at $0.02<z<0.2$ (solid black contour),
%   $0.5<z<1.5$ (dashed red countour), and $1.5<z<2.5$ (dotted purple
%   contour), and from \citet{law12} at $2.5<z<3.6$ (dash-dotted blue
%   contour).

%\subsection{Optical Spectroscopy}
\subsection{Outflow Kinematics and Covering Factors}

We present $\lrest=2500$--5200~\AA\ spectroscopy for three
high-$\sigmasfr$ sources in Figure~\ref{fig:spectra} based on data
from MMT/Blue Channel and SDSS (J1506+5402), Magellan/MAGE
(J1341-0321), and Keck/LRIS and Keck/HIRES (J0905+5759).  The spectra
are dominated by light from the young ($t<50$~Myr) stellar population.
We highlight the \mgii~$\lambda\lambda2796,2803$ absorption lines,
which are used used to measure outflow velocities.  At low spectral
resolution (e.g., the top right panel of Figure~\ref{fig:spectra}) it
is not possible to determine the intrinsic shape of the absorption
line profile and therefore the covering factor of the outflowing gas.
However, the Keck/HIRES spectrum of J0905+5759
($\textnormal{FWHM}\approx8$~\kms) reveals that the gas covers the
entire continuum source near the velocity centroid ($v=-2470$~\kms)
indicating a galaxy-wide outflow.

% MMT/Blue Channel (J1506+5402, $R\sim1000$), SDSS (J1506+5402,
% $R\sim2000$), Magellan/MAGE (J1341-0321, $R\sim4000$), Keck/LRIS
% (J0905+5759, $R\sim600$), and Keck/HIRES (J0905+5759, $R\sim37,000$)

@@

%\section{Discussion}

\subsection{Constraints on Ongoing AGN Activity}

While the SEDs (Figure~\ref{fig:seds}) and optical spectra
(Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the level of ongoing AGN activity and its effect on our
measurements.  Among the high-$\sigmasfr$ galaxies, the strongest case
for detectable AGN emission can be made for J1506+5402, which has the
most luminous \oiii~$\lambda5007$ line among the 29 galaxies observed
by HST.  This source also has a weak \nev~$\lambda3426$ emission line,
which is associated with AGN activity \citep[e.g.,][]{abe08,gil10}.
It was observed with the Chandra X-ray Observatory (proposal ID
11700896, PI: C. Tremonti) and had four detected counts, corresponding
to a 0.5--8.0~keV X-ray luminosity of $10^{42.7}$~erg~s$^{-1}$ (see
P. Sell et al., in preparation for more details on the 12/29 galaxies
with Chandra observations).  Using the relationship between 2--10~keV
X-ray and 12.3~$\mu$m mid-IR luminosity for AGNs from \citet{gan09},
one would expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than observed luminosity for J1506+5402
($L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$).
Based on its \oiii\ luminosity ($10^{42.1}$~erg~s$^{-1}$, a
significant fraction of which could be attributed to star formation)
and the calibration for type 1 AGNs from \citet{hec05}, one would
expect an intrinsic 2--10~keV luminosity of
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{43}$~erg~s$^{-1}$,
suggesting absorption by a factor of $\sim10$ in the X-rays.  However,
even if the X-ray attenuation were a factor of 100, which is the
typical value for local Compton-thick AGNs \citep{dia09}, the expected
mid-IR AGN contribution would only be relevant at the $\lesssim$30\%
level.  Thus, similar to local ULIRGs \citep[e.g.,][]{far07}, the
bolometric output of the galaxies in our sample is dominated by star
formation, and we conclude that our results regarding large
$\sigmasfr$ values are not affected by AGN contamination.

% ased on its \oiii\ luminosity $L(\oiii)=10^{42.1}$~erg~s$^{-1}$, the
% expected X-ray luminosity \citep[adopting the relationship between
%   \oiii\ and X-ray luminosity for type 1 AGNs from][]{hec05}

% No evidence in spectra for unobscured AGN component.  J1506+5402 has
% X-ray and \nev.  Agreement between SFR indicators.  Mention Balmer
% lines and dust geometry.  SEDs well fit by star-forming tempates.
% Most ULIRGs are starburst dominated.  Mention Pauls's paper.  There is
% likely some ongoing AGN activity in sources like J1506+5402, but in
% general, the AGN does not seem to be relevant energetically.  Point
% that high-velocity outflows don't mean that an outflow is AGN driven.

\subsection{The Outflow Launching Mechanism}

It is worth considering whether the high-velocity outflows we observe
were produced by the compact starburst.  \citet{mur11} argued that
massive star clusters with large gas surface densities are the ideal
launching point for galactic-scale outflows driven by radiation
pressure, and that the outflow velocity should scale with escape
velocity of the most massive star clusters in a galaxy.  The
$\gtrsim1000$~\kms\ outflow velocities we observe correspond
approximately to the escape velocities for a compact stellar
population with the sizes and masses that we've measured.
\begin{eqnarray}\nonumber
v_{esc}&=&\sqrt{2GM_{*}/r} \\
&=& 2100
\left({M_{*}\over10^{11} M_{\odot}}\right)^{1/2} 
\left({r\over200~\textnormal{pc}}\right)^{-1/2}
\textnormal{km~s}^{-1} 
\label{eq:vesc}
\end{eqnarray}
This argument, combined with the fact that we observe galaxies with
significant dust-obscured star formation and $\sigmasfr$ values near
the Eddington limit, indicates that momentum input from massive stars
in the form of radiation pressure is a viable mechanism for launching
these outflows.

In addition to radiation pressure, we also expect significant momentum
flux from stellar winds and supernovae.  For example, a starburst with
$\textnormal{SFR}\approx200$~\msun~yr$^{-1}$ would be associated with
radiation pressure $L_{bol}/c\approx2\times10^{35}$~dyne and ram
pressure $\dot{p}\approx5$--$10\times10^{35}$~dyne from stellar winds
and supernovae \citep[e.g.,][]{lei92,hec93,lei99,vei05}.  Furthermore,
\citet{hec11} noted that such a large momentum injection
($\dot{p}\approx10^{35}$~dyne) from a small initial radius
$r_{0}\approx100$~pc could accelerate a cloud with column density
$N_{H}\approx10^{21}$~cm$^{-2}$ to a terminal velocity
$v_{\infty}\approx1800$~\kms.  Thus, the energetics of compact
starbursts are sufficient to produce the high-velocity outflows we
observe, and it is plausible that both radiation pressure on dust
grains and supernova ram pressure contribute to driving the winds.

% Following \citet{hec93} and
% \citet{hec11}, the expected terminal velocity for a cloud with column
% density $\approx10^{21}$~cm$^{-2}$ launched from a radius
% $\approx200$~pc and driven by momentum flux
% $\approx2\times10^{35}$~dynes is:
% \begin{eqnarray}\nonumber
% v_{\infty}&=&1800~\textnormal{km~s}^{-1}
% \left(\dot{p}\over2\times10^{35}~\textnormal{dyne}\right)^{1/2} \\
% &&
% \left({r\over200~\textnormal{pc}}\right)^{-1/2}
% \left({N\over10^{21}~\textnormal{cm}^{-2}}\right)^{-1/2}
% \label{eq:vterm}
% \end{eqnarray}
% The expected momentum flux from stellar winds and supernovae for a
% starburst with $\textnormal{SFR}\approx200$~\msun~yr$^{-1}$ is
% $\approx5$--$10\times10^{35}$~dynes \citep{lei92,hec93,lei99,vei05}
% while the $L_{bol}/c$ radiation pressure is
% $\approx2\times10^{35}$~dynes.  Thus, there is more than enough
% momentum available from stellar winds and supernovae or from radiation
% pressure to produce the outflow velocities we observe.

%AGN have been invoked to launch high-velocity winds.

% \subsection{Momentum Flux from Supernovae and Stellar Winds}
% 
% \citet{meu97} argued for a limit near 45~\units (somewhat smaller for
% a Chabrier IMF) based on the surface brightness of starbursts measured
% using rest-frame UV, H$\alpha$, far-IR, and radio continuum emission,
% such that only 10\% of starbursts exceeded this limit.  They argued
% that the momentum flux from supernovae and stellar winds expected at
% this threshold matches the mean central pressure measured by
% \citet{hec90} for starburst-driven winds.
% 
% There are also relevant arguments about ram pressure driving of
% high-velocity winds from \citet{hec11}.
% 
% 
% \subsection{The Eddington Limit from Radiation Pressure on Dust Grains}

% In models where gaseous outflows are driven by radiation pressure from
% massive stars, estimates of the maximum $\sigmasfr$ where the outward
% force from radiation exceeds the inward force of gravity are in the
% $\sigmasfr\sim10^3$~\units\ range \citep[e.g.,][]{tho05,hop10}.

@@

% The relative importance of momentum injection versus thermal heating
% may vary as a function of galaxy mass \citep[e.g.,][]{hop12}, with
% thermal heating being less efficient in galaxies with larger gas
% densities and shorter cooling times.  Regarding momentum injection,
% \citet{sha11} have argued that radiation pressure becomes more
% important than ram pressure at large SFRs.

%\subsection{Outflows at the Escape Velocity}

@@

High-velocity outflows do not imply AGN feedback.

Theoretical arguments about ram v. radiation pressure.  Point that we
are observing 10$^4$~K gas.

Sample as a whole. Interesting that high-$\sigmasfr$ galaxies tend to
have larger outflow velocities.

\section{Conclusions}

@@

\bibitem[Farrah et al.(2007)]{far07} Farrah, D., Bernard-Salas, J.,
  Spoon, H.~W.~W., et al.\ 2007, \apj, 667, 149

@@

\bibitem[Gandhi et al.(2009)]{gan09} Gandhi, P., Horst, H., Smette,
  A., et al.\ 2009, \aap, 502, 457
\bibitem[Gilli et al.(2010)]{gil10} Gilli, R., Vignali, C., Mignoli,
  M., et al.\ 2010, \aap, 519, A92

@@

\bibitem[Heckman et al.(1993)]{hec93} Heckman, T.~M., Lehnert, M.~D.,
  \& Armus, L.\ 1993, The Environment and Evolution of Galaxies, 188,
  455

@@

\bibitem[Leitherer et al.(1999)]{lei99} Leitherer, C., Schaerer, D.,
  Goldader, J.~D., et al.\ 1999, \apjs, 123, 3

@@

\bibitem[Veilleux et al.(2005)]{vei05} Veilleux, S., Cecil, G., \&
  Bland-Hawthorn, J.\ 2005, \araa, 43, 769
